---
title: "Climate Data Analysis"
date: '2017-10-31T22:42:51-05:00'
description: It has been a pleasure working with you
draft: no
image: null
keywords: ''
slug: project3
categories:
- ''
- ''
---
# Climate change and temperature anomalies

If we wanted to study climate change, we can find data on the *Combined Land-Surface Air and Sea-Surface Water Temperature Anomalies* in the Northern Hemisphere at [NASA's Goddard Institute for Space Studies](https://data.giss.nasa.gov/gistemp). The [tabular data of temperature anomalies can be found here](https://data.giss.nasa.gov/gistemp/tabledata_v4/NH.Ts+dSST.txt).

The period between 1951 - 1980 is our reference in defining temperature anomalies.

```{r load-libraries, include=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
library(scales)
library(hms)
library(infer)
```

## Loading the data set

```{r weather_data, cache=TRUE}

weather <- 
  read_csv("https://data.giss.nasa.gov/gistemp/tabledata_v4/NH.Ts+dSST.csv", 
           skip = 1,  #real data table only starts in Row 2, so we need to skip one row. 
           na = "***") 

```

For each month and year, the data frame shows the deviation of temperature from the normal (expected).

## Isolating the columns we need and tidy data

```{r tidyweather}

tidyweather <- weather %>% 
  select(c("Year", "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) %>%
  pivot_longer(cols=c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec'),
               names_to = "Month", values_to = "Delta") 

```

## Plotting information

```{r scatter_plot}

tidyweather <- tidyweather %>%
  mutate(date = ymd(paste(as.character(Year), Month, "1")),
         Month = month(date, label=TRUE),
         Year = year(date))

```

```{r graphDeltaWeather, fig.width=11}
ggplot(tidyweather, aes(x=date, y=Delta))+
  geom_point(size = 0.5)+
  geom_smooth(color="red") +
  theme_bw() +
  labs (
    title = "Weather Anomalies",
    subtitle = "Temperature progression over time",
    x = "Year",
    y = "Delta in T (°C)"
  )+
  # scale_x_date(date_breaks = seq(min(tidyweather$date), max(tidyweather$date), by = 8300))
     scale_x_date(date_labels = "%Y", date_breaks = "10 years")

```

Scatter plot for each month:

```{r facet_wrap, fig.width = 11}

tidyweather$Month <- factor(tidyweather$Month, 
                            levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) #so that the facet wrap shows according to months

ggplot(tidyweather, aes(x=date, y=Delta))+
  geom_point(size = 0.3)+
  geom_smooth(color="red") +
  theme_bw() +
  facet_wrap(vars(Month)) +
  labs (
     title = "Weather Anomalies broken down by months",
    subtitle = "Temperature progression over time",
    x = "Year",
    y = "Delta in T (°C)"
  )

```

*Based on our analysis, the effect of increasing temperature seems to be constant throughout the year. A more obvious indicator would be the Year instead - the effect of increasing temperature is more pronounced after the period between 1951-1980 as we can see a steep increase in the graph.*

## Grouping the data into time periods

```{r intervals}

comparison <- tidyweather %>% 
  filter(Year>= 1881) %>%     #remove years prior to 1881
  #create new variable 'interval', and assign values based on criteria below:
  mutate(Interval = case_when(
    Year %in% c(1881:1920) ~ "1881-1920",
    Year %in% c(1921:1950) ~ "1921-1950",
    Year %in% c(1951:1980) ~ "1951-1980",
    Year %in% c(1981:2010) ~ "1981-2010",
    TRUE ~ "2011-present"
  ))

```

## Distribution of monthly deviations

```{r density_plot, fig.width=11}

ggplot(comparison, aes(x=Delta, fill=Interval))+
  geom_density(alpha=0.2) +   #density plot with transparency set to 20%
  theme_bw() +                #theme
  labs (
    title = "Density Plot for Monthly Temperature Anomalies",
    y     = "Density",         #changing y-axis label to sentence case
    x = "Delta in T (°C)"
  )

```

## Average annual anomalies

```{r averaging, fig.width=11}

#creating yearly averages
average_annual_anomaly <- tidyweather %>% 
  group_by(Year) %>%   
  summarise(annual_average_delta = mean(Delta, na.rm=TRUE))

#plotting the data:
ggplot(average_annual_anomaly, aes(x=Year, y= annual_average_delta))+
  geom_point(size = 0.8)+
  geom_smooth(colour = "red") +
  theme_bw() +
labs (
     title = "Average Yearly Temperature Anomaly",
    subtitle = "Temperature progression over time",
    x = "Year",
    y = "Delta in T (°C)"
  )+
  scale_x_continuous(n.breaks = 20)



```

## Confidence Interval for `delta`

[NASA points out on their website](https://earthobservatory.nasa.gov/world-of-change/decadaltemp.php) that

> A one-degree global change is significant because it takes a vast amount of heat to warm all the oceans, atmosphere, and land by that much. In the past, a one- to two-degree drop was all it took to plunge the Earth into the Little Ice Age.

### Confidence Interval for `delta` (Formula)

```{r calculate_CI_using_formula}

formula_CI <- comparison %>% 
  filter(Interval == "2011-present") %>%
  summarize(median_delta = median(Delta, na.rm=TRUE),
            mean_delta = mean(Delta, na.rm=TRUE), 
            sd_delta = sd(Delta, na.rm=TRUE), 
            count = n(), 
            t_critical = qt(0.975, count -1),
            se_delta = sd_delta/sqrt(count), 
            margin_of_error = t_critical * se_delta,
            lower_CI_delta = mean_delta - margin_of_error,
            upper_CI_delta = mean_delta + margin_of_error
  )
            
#print out formula_CI
formula_CI

```

### Confidence Interval for `delta` (Bootstrapping)

```{r calculate_CI_using_bootstrap}

library(infer)

set.seed(1234)

boot_delta <- comparison %>% 
  filter(Interval == "2011-present") %>% 
  specify(response = Delta) %>% 
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat="mean") 

percentile_ci <- boot_delta %>%  
  get_ci(level = 0.95, type = "percentile")

#print out bootstrap_CI
percentile_ci

```

### Comparison of Confidence Intervals

```{r visualizing_bootstrap, fig.height=9, fig.width=11}

visualize(boot_delta) +
  shade_ci(endpoints = percentile_ci) + 
  geom_vline(xintercept = formula_CI$lower_CI_delta, linetype = "dashed", colour = "red", size = 1) +
  geom_vline(xintercept = formula_CI$upper_CI_delta,linetype = "dashed", colour = "red", size = 1) + 
  theme_bw() +
  labs(
    title = "Bootstrap CI for Average Annual Delta since 2011", 
    subtitle = "Fomula CI shown with dashed red lines",
    x = "Delta T (°C)",
    y = "Count"
    
  )

```

*Based on our analysis, the 95% confidence interval (CI) for the average annual delta since 2011 was calculated as (1.01, 1.11). This means that we can be 95% confident that the average annual delta falls within 1.01 and 1.11. We constructed the CI using the formula method and using a bootstrap simulation, both method gives us a very similar width - the shaded area is the CI calculated using a bootstrap simulation and the range between the dotted lines is the CI calculated using the formula.*